---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(squarebrackets)
set.seed(1)
```

# squarebrackets

<!-- badges: start -->
`r badge_repostatus("WIP")`
`r badge_lifecycle("experimental")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



squarebrackets: Methods as an Alternative to the Square Brackets Operators

# Description

Provides subset methods (supporting both non-recursive and recursive S3 classes) that may be more convenient alternatives to the `[` and `[<-` operators, whilst maintaining similar performance.

Some nice properties of these methods include, but are not limited to, the following:

  1) The`[` and `[<-` operators use different rulesets for data.frame-like types (data.frames, tibbles, data.tables, tibbles, etc.). The 'squarebrackets' methods use consistent methods for these data.frame-like types.
  2) Performing dimensional subset operations on an array using `[` and `[<-`, requires a-priori knowledge of the number of dimensions the array has. The 'squarebrackets' methods work on any arbitrary dimensions without requiring such prior knowledge.
  3) When selecting names with the `[` and `[<-` operators, only the first occurrence of the names are selected in case of duplicate names. The 'squarebrackets' methods always perform on all names in case of duplicates, not just the first.
  4) The `[<-` operator only supports copy-on-modify semantics for most classes. The 'squarebrackets' methods provides explicit pass-by-reference and pass-by-value semantics, whilst still respecting things like binding-locks and mutability rules.

&nbsp;


## Get Started

To get started see `?squarebrackets_help`


&nbsp;


## Installing & Loading

One can install 'squarebrackets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/squarebrackets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(squarebrackets)
```


Should the user wish to expose specific functions from 'squarebrackets' only within the current/local environment, one can use the following:


```{r eval = FALSE, echo = TRUE}
tinycodet::import_LL("squarebrackets", selection = ... )
```

&nbsp;


## Changelog (EXPERIMENTAL VERSIONS)

 - 10 March 2024: First GitHub upload - Package is very much experimental.
 - 12 March 2024: Changed the introduction help page a bit, added `dt_setadd()`, and added tests for all `dt_` - functions. There are slightly over 50,000 tests now.
 - 15 March 2024: Added the `sb_setRename()` method, and added tests for this method also.
 - 16 March 2024: Fixed a bug in the "rcpp_set_rowcol" source code. Tweaked the documentation here and there a bit. Improved the tests a bit.
 - 17 March 2024: Added more tests, and tweaked the documentation a bit.
 - 19 March 2024: Some methods/functions did not support mutable_atomic type "complex"; this is now fixed. Added support for mutable_atomic type "raw". Added tests for atomic type handling. Added the functions `ma_setv()`, and `couldb.mutable_atomic()`. Added the options "sb.rat" and "sb.chkdup"; argument `chkdup` is now also set to `FALSE` by default. Added more badges to the documentation.
 - 20 March 2024: The user can now also specify `coe = TRUE` in `sb_mod.data.frame()`.
 - 24 March 2024: Methods are now split between methods for non-recursive objects (`sb_`), and methods for recursive objects (`sb2_`).
 - 26 March 2024: Replaced `seq_rec()` with `seq_rec2()`.
 - 27 March 2024: Added `dt_setreorder()`, and added tests for this also. 'abind' now as a dependency, and 'abind' based code removed, as it is redundant.
 - 29 March 2024: Added `sb2_before.array()` and `sb2_after.array()`, and added tests for these also. Added tests for data.frame-like coercion types. Tweaked the documentation here and there a bit.
 - 30 March 2024: Removed the separate `NA` checks, as they are redundant. Fixed some linguistic mistakes in the documentation.
 - 1 April 2024: Removed `sb_coe()` but kept `sb2_coe()`. Added `inv` argument to `sb_mod()`/`sb2_mod()`, `sb_set()`/`sb2_set()`, and `sb2_coe()`, and added tests for these. Added `idx1()` for Copy-On-Modification Substitution, and added tests for `idx1()` also.Fixed a bug in character subset ordering in the `sb`/`sb2`_`mod`/`set`/`coe` - generic methods. Fixed a bug in the introduction message. Added even more tests. Added `idx1_dim()`, and added tests for these also.
 - 5 April 2024: Replaced `idx1`/`idx0` with `idx()`.
 - 18 May 2024: Added a few tests for the `idx()` method (need to add more). Fixed the export pattern expressions in the Namespace file. Adjusted the documentation.
 

&nbsp;
