---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(squarebrackets)
set.seed(1)
```

# squarebrackets

<!-- badges: start -->
`r badge_github_actions("tony-aw/squarebrackets")`
`r badge_repostatus("WIP")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



squarebrackets: Subset Methods as Alternatives to the Square Brackets Operators for Programming


## What is the 'squarebrackets' R package?

'squarebrackets' is a light-weight extension of the base 'R' `[`, `[<-`, `[[`, and `[[<-` operators.

&nbsp;

## Why use 'squarebrackets'?

```{r echo = FALSE, eval = TRUE, results = 'asis'}

txt <- packageDescription("squarebrackets", fields = "Description")
p <- c("\t", ",\n", ".\n", ";\n",  "\n(", "following.")
rp <- c("", ", ", ".  \n\n", "; \n", " (", "following:")
for(i in 1:length(rp)) {
 txt <- gsub(p[i], rp[i], txt, fixed = TRUE)
}
txt <- paste0(txt, "\n \n")
txt <- gsub("(?m)^[ ]{4,}", "", txt, perl = TRUE)

cat(txt)

```


&nbsp;


## Get Started

To get started see the [Get Started](https://tony-aw.github.io/squarebrackets/articles/squarebrackets.html) page.


&nbsp;


## Installing & Loading

One can install 'squarebrackets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/squarebrackets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(squarebrackets)
```


&nbsp;

&nbsp;
