---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(squarebrackets)
set.seed(1)
```

# squarebrackets

<!-- badges: start -->
`r badge_repostatus("WIP")`
`r badge_lifecycle("experimental")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



squarebrackets: Subset Methods as Alternatives to the Square Brackets Operators for Programming

## Description

Provides subset methods (supporting both atomic and recursive S3 classes) that may be more convenient alternatives to the `[` and `[<-` operators, whilst maintaining similar performance.

Some nice properties of these methods include, but are not limited to, the following:

  1) The `[` and `[<-` operators use different rule-sets for different data.frame-like types (data.frames, data.tables, tibbles, tidytables, etc.). The 'squarebrackets' methods use the same rule-sets for the different data.frame-like types.
  2) Performing dimensional subset operations on an array using `[` and `[<-`, requires a-priori knowledge on the number of dimensions the array has. The 'squarebrackets' methods work on any arbitrary dimensions without requiring such prior knowledge.
  3) When selecting names with the `[` and `[<-` operators, only the first occurrence of the names are selected in case of duplicate names. The 'squarebrackets' methods always perform on all names in case of duplicates, not just the first.
  4) The `[[` and `[[<-` operators allow operating on a recursive subset of a nested list. But these only operate on a single recursive subset, and are not vectorized for multiple recursive subsets of a nested list at once. 'squarebrackets' provides a way to reshape a nested list into a recursive matrix, thereby allowing vectorized operations on recursive subsets of such a nested list.
  5) The `[<-` operator only supports copy-on-modify semantics for most classes. The 'squarebrackets' methods provides explicit pass-by-reference and pass-by-value semantics, whilst still respecting things like binding-locks and mutability rules.
  6) 'squarebrackets' supports index-less sub-set operations, which is more memory efficient (and better for the environment) for `long vectors` than sub-set operations using the `[` and `[<-` operators.

&nbsp;


## Get Started

To get started see `?squarebrackets_help`


&nbsp;


## Installing & Loading

One can install 'squarebrackets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/squarebrackets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(squarebrackets)
```


&nbsp;


## Changelog (EXPERIMENTAL VERSIONS)

 - 10 March 2024: First GitHub upload - Package is very much experimental.
 - 12 March 2024: Changed the introduction help page a bit, added `dt_setadd()`, and added tests for all `dt_` - functions. There are slightly over 50,000 tests now.
 - 15 March 2024: Added the `sb_setRename()` method, and added tests for this method also.
 - 16 March 2024: Fixed a bug in the "rcpp_set_rowcol" source code. Tweaked the documentation here and there a bit. Improved the tests a bit.
 - 17 March 2024: Added more tests, and tweaked the documentation a bit.
 - 19 March 2024: Some methods/functions did not support mutable_atomic type "complex"; this is now fixed. Added support for mutable_atomic type "raw". Added tests for atomic type handling. Added the functions `ma_setv()`, and `couldb.mutable_atomic()`. Added the options "sb.rat" and "sb.chkdup"; argument `chkdup` is now also set to `FALSE` by default. Added more badges to the documentation.
 - 20 March 2024: The user can now also specify `coe = TRUE` in `sb_mod.data.frame()`.
 - 24 March 2024: Methods are now split between methods for non-recursive objects (`sb_`), and methods for recursive objects (`sb2_`).
 - 26 March 2024: Replaced `seq_rec()` with `seq_rec2()`.
 - 27 March 2024: Added `dt_setreorder()`, and added tests for this also. 'abind' now as a dependency, and 'abind' based code removed, as it is redundant.
 - 29 March 2024: Added `sb2_before.array()` and `sb2_after.array()`, and added tests for these also. Added tests for data.frame-like coercion types. Tweaked the documentation here and there a bit.
 - 30 March 2024: Removed the separate `NA` checks, as they are redundant. Fixed some linguistic mistakes in the documentation.
 - 1 April 2024: Removed `sb_coe()` but kept `sb2_coe()`. Added `inv` argument to `sb_mod()`/`sb2_mod()`, `sb_set()`/`sb2_set()`, and `sb2_coe()`, and added tests for these. Added `idx1()` for Copy-On-Modification Substitution, and added tests for `idx1()` also.Fixed a bug in character subset ordering in the `sb`/`sb2`_`mod`/`set`/`coe` - generic methods. Fixed a bug in the introduction message. Added even more tests. Added `idx1_dim()`, and added tests for these also.
 - 5 April 2024: Replaced `idx1`/`idx0` with `idx()`.
 - 18 May 2024: Added a few tests for the `idx()` method (need to add more). Fixed the export pattern expressions in the Namespace file. Adjusted the documentation.
 - 26 May 2024: Removed `sb2_coe()`, as it is redundant.
 - 6 June 2024: Removed the `sb(2)_before/after` methods in favour of the the new `bind_`/`bind2_` implementations. Added the `lst_` functions. Added the `options` help page.
 - 30 June 2024: Re-written internal code for arrays. Added support for backward indexing via Complex Vector indices. Added more tests. Replaced `seq_names()` with the new and far more flexible `idx_r()` function.
 - 31 August 2024: Made the tests more efficient. Removed separate method dispatch for factors, as using the default atomic vector method dispatch is sufficient for factors.
 - 7 September 2024: Incorporated some ALTREP functionality into the package.
 - 15 September 2024: Replaced the `drop` argument with `red` to avoid confusion with base R's own `drop` mechanic. Small performance improvements for `sub2ind()` and `sb_set.array()`.
 - 26 September 2024: Overhauled how indexing with complex vectors work.
 - 28 September 2024: Split `sb(2)_setRename()` into `sb_setFlatnames()`, `sb_setDimnames()`, and `sb2_setVarnames()`.
 - 10 October 2024: `sb_mod()` now makes partial copies of data.frame-like objects instead of whole copies, for more memory efficiency. Also removed the old `sb_str()` and `sb_a()` functions. Renamed `ci_seq()` to `cp_seq()` (in preparation for the next update).
 - 19 October 2024: Removed the renaming methods (`sb_setRename`), and `seq_rec2()`. Added `slcseq_`.
 - 5 November 2024: Renamed `slcseq_` to `slice_`. Re-organized the documentation a bit. Fixed examples of the `ci` and `tci` help pages. Added `bind_mat()` and `bind2_mat()`. Added `ndims()`.
 - 14 November 2024: Performance improvement of `match_all()`.
 - 21 November 2024: Improved the documentation. Slightly tweaked array argument usage. Added `sticky` option. Brought back the renaming methods. Changed behaviour of the `use.names` argument in `lst_untree()`.
 - 24 November 2024: Matrices now use the same API as arrays. Adjusted the documentation accordingly. Cleaned up the internal code a bit.
 - 30 November 2024: The binding implementations can now bind mixtures of atomic and recursive objects.
 - 5 December 2024: replaced the `_rm` post-fixes with `_wo` in all methods, to avoid confusion. Coercion for data.frame-like objects now happens automatically, and only when needed, in the `sb2_mod()` method, and updated the documentation accordingly. Slightly re-organized the documentation.



&nbsp;
