---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# squarebrackets

<!-- badges: start -->

[![R-CMD-check-overkill](https://github.com/tony-aw/squarebrackets/actions/workflows/R-CMD-check-overkill.yaml/badge.svg)](https://github.com/tony-aw/squarebrackets/actions/workflows/R-CMD-check-overkill.yaml)
[![Project Status: WIP - Initial development is in progress, but there
has not yet been a stable, usable release suitable for the
public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![](https://img.shields.io/badge/ORCID-0000--0001--9498--8379-green.svg)](https://orcid.org/0000-0001-9498-8379)
<!-- badges: end -->

squarebrackets: Subset Methods as Alternatives to the Square Brackets
Operators for Programming

## What is the ‘squarebrackets’ R package?

‘squarebrackets’ is a light-weight extension of the base ‘R’ `[`, `[<-`,
`[[`, and `[[<-` operators.

&nbsp;


```{r echo = FALSE, eval = TRUE, results = 'asis'}
 
txt <- packageDescription("squarebrackets", fields = "Description")
p <- c("\t", ",\n", ".\n", ";\n",  "\n(", "following.")
rp <- c("", ", ", ".<br><br>", "; <br> ", " (", "following:")
for(i in 1:length(rp)) {
txt <- gsub(p[i], rp[i], txt, fixed = TRUE)
}
txt <- paste0(txt, "\n \n")
cat("<p style='white-space: pre-line;'>", txt, "</p>", sep = "")

```


&nbsp;

## Get Started

To get started see the [Get
Started](https://tony-aw.github.io/squarebrackets/articles/squarebrackets.html)
page.

&nbsp;

## Installing & Loading

One can install ‘squarebrackets’ from GitHub like so:

```{r, eval=FALSE}
remotes::install_github("https://github.com/tony-aw/squarebrackets")
```

Special care has been taken to make sure the function names are clear,
and that the function names are unlikely to conflict with core R, the
recommended R packages, the rstudioapi package, or major packages from
the fastverse. So one can attach the package - thus exposing its
functions to the namespace - using:

```{r, eval=FALSE}
library(squarebrackets)
```


