---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(squarebrackets)
set.seed(1)
```

# squarebrackets

<!-- badges: start -->
`r badge_repostatus("WIP")`
`r badge_lifecycle("experimental")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



squarebrackets: Methods as an Alternative to the Square Brackets Operators

&nbsp;

Among programming languages, 'R' has perhaps one of the most flexible and comprehensive sub-setting functionality, provided by the square brackets operators (`[`, `[ <-`).

But in some occasions, the square brackets operators are less than optimally convenient; examples:

 - Performing subset operations on an array, when the number of dimensions of the array are not known a-priori, is quite convoluted.
 - The `data.frame`, `tibble`, `data.table` and `tidytable` classes use different rule sets for the square brackets operators. So one needs to learn new rules for different data.frame classes, making your one's code inconsistent.
 -  Selecting names for sub-setting, only the first occurrences of the names are selected for the sub-set. And when un-selecting/removing names for sub-setting, the syntax is very different from selecting names
 - 'R' implicitly uses copy-on-modify semantics. The `[<-` operator does not support explicitly choosing between deep copy or pass-by-reference semantics.

The 'squarebrackets' package provides methods as an alternative to the square brackets operators, that are more convenient in such occasions.


To get started see `?squarebrackets_help`


&nbsp;


## Installing & Loading

One can install 'squarebrackets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/squarebrackets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(squarebrackets)
```


Should the user wish to expose specific functions from 'squarebrackets' ONLY within a specific environment, like only within a specific function, one can use the following:


```{r eval = FALSE, echo = TRUE}
tinycodet::import_LL("squarebrackets", selection = ... )
```

&nbsp;


## Changelog (EXPERIMENTAL VERSIONS)

 - 10 March 2024: First GitHub upload - Package is very much experimental.
 - 12 March 2024: Changed the introduction help page a bit, added `dt_setadd()`, and added tests for all `dt_` - functions. There are slightly over 50,000 tests now.
 - 15 March 2024: Added the `sb_setRename()` method, and added tests for this method also.
 - 16 March 2024: Fixed a bug in the "rcpp_set_rowcol" source code. Tweaked the documentation here and there a bit. Improved the tests a bit.
 - 17 March 2024: Added more tests, and tweaked the documentation a bit.
 - 19 March 2024: Some methods/functions did not support mutable_atomic type "complex"; this is now fixed. Added support for mutable_atomic type "raw". Added tests for atomic type handling. Tweaked the documentation a bit. Added the functions `ma_setv()`, and `couldb.mutable_atomic()`.


 

&nbsp;
