---
title: "Benchmarks - sb_x"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Benchmarks - sb_x}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(squarebrackets)
```


&nbsp;

# Introduction

Due to the many checks and conversions performed by the `squarebrackets::` functions, to make sub-setting more programmatically and beginner friendly, the functions are almost necessarily slower than base R's `[`-like operators.

However, a considerable effort was made to keep the speed loss to a minimum. Generally, the speed loss is indeed negligible, and in some cases there is even speed improvement (thanks to the heavy lifting performed by the 'collapse' package).

Below are some benchmarks to give one an idea of the speed loss. These are just examples; speed is determined by a great number of factors.

&nbsp;

```{r}
library(bench)
library(ggplot2)
library(patchwork)
```

```{r echo = FALSE}
plotfun <- function(p1, p2) {
  combined <- p1 + p2 & theme(legend.position = "bottom")
  combined + plot_layout(guides = "collect")
}
```

# Atomic objects

## Matrix

```{r eval=FALSE, echo=TRUE}

n <- 5e3
x.mat <- matrix(seq_len(n*n), ncol = n)
colnames(x.mat) <- sample(c(letters, NA), n, TRUE)
sel.rows <- 1:100
sel.cols <- rep(sample(letters[1:13]), 10)
bm.sb_x.matrix <- bench::mark(
  "sb_x" = sb_x(x.mat, sel.rows, sel.cols),
  "base R" = x.mat[sel.rows, lapply(sel.cols, \(i) which(colnames(x.mat) == i)) |> unlist(), drop = FALSE],
  min_iterations = 500
)
bm.sb_x.matrix
summary(bm.sb_x.matrix)

```


```{r echo = FALSE, eval = TRUE}
load("bm.sb_x.matrix.RData")
summary(bm.sb_x.matrix)
```


&nbsp;

## Array (3D)

```{r eval=FALSE, echo=TRUE}
x.dims <- c(5000, 2000, 4)
x.3d <- array(1:prod(x.dims), x.dims)
sel.rows <- 1:900
sel.lyrs <- c(TRUE, FALSE, TRUE, FALSE)
bm.sb_x.3d <- bench::mark(
  "squarebrackets" =  sb_x(x.3d, n(sel.rows, sel.lyrs), c(1,3)),
  "base R + abind" = abind::asub(x.3d, idx = list(sel.rows, sel.lyrs), dims = c(1,3)),
  min_iterations = 500
)
summary(bm.sb_x.3d)
```

```{r echo = FALSE, eval = TRUE}
load("bm.sb_x.3d.RData")
summary(bm.sb_x.3d)
```

&nbsp;

## Plot

```{r echo = FALSE, fig.width = 8}
p.mat <- autoplot(bm.sb_x.matrix) + ggtitle("matrix")
p.3d <- autoplot(bm.sb_x.3d) + ggtitle("array (3d)")
plotfun(p.mat, p.3d)
```


&nbsp;

# Data.frame-like objects

## data.frame

```{r eval=FALSE, echo=TRUE}
n <- 1e5
chrmat <- matrix(
  sample(letters, n*400, replace = TRUE), ncol = 400
)
intmat <- matrix(
  seq.int(n*400), ncol = 400
)
x <- cbind(chrmat, intmat) |> as.data.frame()
rm(list = c("chrmat", "intmat"))
colnames(x) <- make.names(colnames(x), unique = TRUE)
sel.cols <- rep(sample(names(x), 10), 4)
sel.rows <- 1:1000
bm.sb_x.df <- bench::mark(
  "squarebrackets" = sb2_x(x, sel.rows, sel.cols),
  "collapse::ss" = collapse::ss(x, sel.rows, match(sel.cols, names(x))),
  "base R" = x[sel.rows, match(sel.cols, names(x)), drop = FALSE],
  min_iterations = 1e4
)
summary(bm.sb_x.df)

```

```{r echo = FALSE, eval = TRUE}
load("bm.sb_x.df.RData")
summary(bm.sb_x.df)
```

&nbsp;

## data.table

```{r eval=FALSE, echo=TRUE}
x <- as.data.table(x)
tempfun <- function(x, i, j) {
  x <- collapse::ss(x, i, j, check = TRUE)
  names(x) <- make.names(names(x), unique = TRUE)
  return(x)
}
bm.sb_x.dt <- bench::mark(
  "squarebrackets" = sb2_x(x, sel.rows, sel.cols),
  "data.table + collapse" = tempfun(x, sel.rows, match(sel.cols, names(x))),
  min_iterations = 1e4
)
summary(bm.sb_x.dt)
```

```{r echo = FALSE, eval = TRUE}
load("bm.sb_x.dt.RData")
summary(bm.sb_x.dt)
```

&nbsp;

## plot

```{r echo = FALSE, fig.width = 8}
p.df <- autoplot(bm.sb_x.df) + ggtitle("data.frame")
p.dt <- autoplot(bm.sb_x.dt) + ggtitle("data.table")
plotfun(p.df, p.dt)
```

&nbsp;
