---
title: "Benchmarks - Vectorized Recursive Subset Operations On Nested Lists"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Benchmarks - Vectorized Recursive Subset Operations On Nested Lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(squarebrackets)
```


&nbsp;

# Introduction

Operating on recursive sub-sets of a nested lists requires the `[[` and `[[<-` operators. These operators can only select a single subset at once. Thus, operating on multiple recursive sub-sets requires a for-loop, and is thus not vectorized.

The `lst_untree()` function unnests a tree-like nested list into a recursive matrix, in such a way that recursive sub-set operations can be performed in a vectorized manner.

Below are the benchmarks on operating on the original list, and the list transformed by `lst_untree()`.


```{r fig.width = 8}

x <- lapply(1:1000, \(x) list(
  list(sample(letters), sample(LETTERS)),
  list(sample(month.abb), sample(month.name)),
  list(sample(1:100), rnorm(100))
))
y <- lst_untree(x, 1)

foo <- bench::mark(
  original = for(i in seq_along(x)) {x[[c(i, c(1,2))]]},
  "untree'd" = sb2_x(n(seq_len(nrow(y)), 2), 1:2),
  check = FALSE,
  min_iterations = 1000
)
summary(foo)
ggplot2::autoplot(foo)

```

&nbsp;
