---
title: "Additional Examples"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Additional Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(squarebrackets)
```

&nbsp;

# Introduction

The function documentation of the 'squarebrackets' package is rich with examples. This article adds additional examples focussed on combining 'squarebrackets' with other packages.

&nbsp;

# Aggregate sf-data.table

The 'sf' package adds data.frames/data.tables with spatial features. In this example, the functionality from 'squarebrackets' is used with an 'sf' class data.table.
Here a plot is created (using 'ggplot2', 'ggh4x', and 'tinycodet') from an aggregation by region of the spatial data.table:

```{r fig.width=8}
library(ggplot2)
library(ggh4x)
library(tinycodet)
import_as(~dt., "data.table")
x <- sf::st_read(system.file("shape/nc.shp", package = "sf"))
x <- dt.$as.data.table(x)

x$region <- ifelse(x$CNTY_ID <= 2000, 'high', 'low')
plotdat <- dt_aggregate(
  x, SDcols = "geometry", f = sf::st_union, by = "region"
)
geometry <- form(~geometry)
fill <- form(~region)
fct <- form(~region)
ggplot(plotdat, aes_pro(geometry = geometry, fill = fill)) + 
  facet_wrap2(fct, nrow = 2) +
  geom_sf()
```


&nbsp;

# Re-order an array like a Rubik's cube

The the `%row~%` and `%col~%` operators from 'tinycodet' can quickly re-order elements of each row or column independently of other rows or columns.

One can combine these operators with the `sb_` methods from 'squarebrackets' to re-order matrix slices of an array, like re-ordering the panels on a Rubik's cube. By using `sb_set()` one can even do so by reference for better memory efficiency - which is exactly what will be done here.

Let's first create a mutable array:

```{r}
x <- mutable_atomic(sample(1:27), dim = c(3,3,3))
print(x)
```

Now let's numerically sort every column in all slices in this array by reference:

```{r}
library(tinycodet)

for(i in 1:3) {
  y <- sb_x(x, n(i), 3) |> drop()
  rp <- y %col~% y
  sb_set(x, n(i), 3, rp = rp)
}
```

Now, let's see if this was successful:

```{r}
print(x)
```

Yes, each column in each layer has independently been sorted numerically.

One can do the same thing with a character array; note that this time we use `stringi::stri_rank()` for determining the proper order:

```{r}
x <- mutable_atomic(sample(c(letters, 1e3)), dim = c(3,3,3))
print(x)
for(i in 1:3) {
  y <- sb_x(x, n(i), 3) |> drop()
  rp <- y %col~% stringi::stri_rank(y)
  sb_set(x, n(i), 3, rp = rp)
}
print(x)
```

Again: successful.


&nbsp;


&nbsp;

# Mutate column of data.table

We can mutate a column of a data.table by reference, by combining 'squarebrackets'  with `tinycodet::with_pro()`, as follows:

```{r}

x <- sf::st_read(system.file("shape/nc.shp", package = "sf"))
x <- dt.$as.data.table(x)

myform <- form(~ PERIMETER / AREA)
x$newcolumn <- 1:nrow(x)
mutated <- dt.$as.data.table(with_pro(x, myform))
sb_set(x, col = "newcolumn", rp = mutated)
summary(x)

```


&nbsp;
